---
title: "What you can do with RMarkdown"
subtitle: "BSTA504"
author: "Ted Laderas"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(xaringanExtra)
library(flair)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "midnightblue",
  header_font_google = google_font("Josefin Sans"),
  link_color = "steelblue",
  text_font_size = "26px",
  footnote_font_size = "22px"
)
```


class: center, middle
# Warning: RMarkdown is a Deep (but Rewarding) Rabbit Hole

---
class: center, middle
# What is RMarkdown?

---
# One format to Rule them All

--
- Variant of Markdown (made for websites)

--
- Lets you mix code and text

--
- Can be made into many different formats

---
# Parts of an RMarkdown Document

- Metadata (front matter)
  - lets you set formatting options, document type
- Text 
  - Uses [Markdown Formatting](https://sph-r-programming.netlify.app/reference/markdown/)
  - Can mix in HTML tags as well (such as twitter embed code)
- Code Chunks
  - Executes code and generates output in order, from scratch

---
# How does this work?

- knitting: generating outputs (tables, figures) from codeblocks
- Output + markdown is then passed down to a "rendering engine" which makes the document
  - Pandoc (html, word documents, EPUB, many others)
  - reveal.js (slide presentations)
  - LaTeX (PDFs)
- Can save intermediate outputs as well.
  - LaTeX code

---
class: center, middle

# Let's learn about Front Matter/Metadata

---
# Front Matter: YAML

You can include a special section at the top of a Markdown document that contains metadata (or data about your document) like the `title`, `date`, `author`, etc. 

This section uses a special syntax named YAML (or “YAML Ain’t Markup Language”) that follows this basic outline: 

```
---                   #note it has to start with three dashes
setting: value for setting. 
---                   #and end in three dashes.
```
---
# Front Matter Example

.pull-left[
```{r yaml_ex, eval=FALSE, include=FALSE}
---
title: "Introduction to RMarkdown"
author: "Ted Laderas"
date: "\`r Sys.Date()\`"
output: 
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    number_sections: true
---

```

```{r echo=FALSE}
decorate_chunk("yaml_ex", eval=FALSE)
```
]

.pull-right[
Here’s an example YAML metadata section. Note that it must start and end with three dashes (---).

]
---
# `title` and `author`

.pull-left[
```{r echo=FALSE}
decorate_chunk("yaml_ex", eval=FALSE) %>%
  flair("title") %>%
  flair("author")
```
]

.pull-right[
- `title` - title of your document
- `author` - author of your document

]
---
# `date`

.pull-left[
```{r echo=FALSE}
decorate_chunk("yaml_ex", eval=FALSE) %>%
  flair("date") %>%
  flair("`r Sys.Date()`")
```
]

.pull-right[
- `date` can be in a regular format, such as `08/20/2012`
- using the inline code chunk updates the date automatically!
]
---
# `output`

.pull-left[

```{r echo=FALSE}
decorate_chunk("yaml_ex", eval=FALSE) %>%
  flair("output:") %>%
  flair("html_document:")
```
]

.pull-right[
- `output:` is one of the most important options to set.
- You have seen `output: html_document` before.
- This is an example of a *nested* option, where an option is part of a larger option.
- Use **two** spaces for the **first** level. Use **four** spaces for the **second** level, and so on.
]
---
# Let's look at three output types

.pull-left[
- `html_document` 
- `distill::distill_article`
- `slidy_presentation`
]

.pull-right[
- Many, many other output formats
- `{xaringan}` presentations (like this)
- Shiny web apps
- LearnR tutorials
- Book format (`{bookdown}`)
- Websites (`{blogdown}`/`{distill}`)

]
---
# `html_document`: a really versatile format!

Lots of options make it a really versatile format.

- Table of contents 
- Themes
- Section Numbering
- Figure Formatting
- Embed interactive visualizations
- Data Frame Printing

---
# Themes

.pull-left[
```{r theme_chunk, include=FALSE, eval=FALSE}
---
output:
  html_document:
    theme: flatly
---
```

```{r, echo=FALSE}
decorate_chunk("theme_chunk", eval=FALSE) %>%
  flair("theme: flatly")
```
]

.pull-right[
There are a number of themes that exist within Rmarkdown. 

Examples include: 
- `flatly` 
- `cosmo` 
- `lumen` 
- `paper`

More info here: [RMarkdown Theme Gallery](https://www.datadreaming.org/post/r-markdown-theme-gallery/)
]
---
# Theme packages that are really helpful

- `{prettydoc}`: https://prettydoc.statr.me/themes.html
- `{hrbrthemes}`: https://cinc.rud.is/web/packages/hrbrthemes/
- `{rmdformats}`: https://bookdown.org/yihui/rmarkdown/rmdformats.html 
- `{tufte}`: https://bookdown.org/yihui/rmarkdown/tufte-handouts.html
- [Branding and Packaging Reports with RMarkdown](https://www.wjakethompson.com/talk/2020-01-30-rstudioconf-ratlas/)

---
# `toc: true`

.pull-left[
```{r echo=FALSE}
decorate_chunk("yaml_ex", eval=FALSE) %>%
  flair("toc: true") %>%
  flair("toc_float: true")
```
]

.pull-right[
- Add a table of contents to your file with `toc: true`
- Setting `toc_float: true` adds a *floating* table of contents
]
---
# number_sections

.pull-left[
```{r num_chunk, include=FALSE, eval=FALSE}
---
output:
  html_document:
    number_sections: true
---
```

```{r echo=FALSE}
decorate_chunk("yaml_ex", eval=FALSE) %>%
  flair("number_sections: true") 
```
]

.pull-right[
- Auto-number each section and heading
]

---

class: center, middle

# Code Chunks, Tables, and Figures

---
# Some best practices

Nick Tierney suggests that you always have these 3 chunks in your file

```
{r setup}
```

Note: this chunk is always run, even if you hit play on a later chunk


```
{r library}
```

Put your library statements here


```
{r functions}
```

Put functions here

.footnote[https://rmd4sci.njtierney.com/using-rmarkdown.html#nicks-rmarkdown-hygiene-recommendations]
---
# Code Chunks and Languages

.pull-left[
```
{r}
```

```
{python}
```
]
.pull-right[
- Doesn't have to just be R! Can be `python` as well
- Other languages are supported, including SQL for databases
]
---
# Code Chunks can have names

```
{r my_code_chunk, fig.height=8, fig.width=10}
```

- add a name right after the `r` (no commas after `r`)
- can be referred to later (useful with autonumbering and referring to figures)

---
# Important Options: `echo=FALSE`

Show my output, but not my code.

```
{r my_code_chunk, echo=FALSE}
```

---
# Important Options: `eval=FALSE`

Show my code block, but don't run it.

```
{r my_code_chunk}
```

---
# Important Options: `warning, message`

If an operation is throwing warnings or giving you an unwanted message, you can set these to be `FALSE`.

```
{r my_code_chunk, warning=FALSE, message=FALSE}
```

---
# `fig.height` and `fig.width`

- Lets you set dimensions of your figures
- Default measurement is in Inches
- Much more here: https://rmd4sci.njtierney.com/customising-your-figures.html

```
{r my_code_chunk, fig.height=8, fig.width=10}
```

---
# Table options

We'll take a look at this in the notebooks.

```
output:
  html_document:
    df_print: paged
```

- `paged`
- `kable`

---
# Setting default chunk options with `knitr::opts_chunk$set()`

Can set the default chunk options with the following in your `setup` chunk:

```
knitr::opts_chunk$set(echo = TRUE, fig.width=8, fig.height=8)
```
---
# Netlify Drop Demo

Go to https://app.netlify.com/drop


---
# Table and figure references

See RMarkdown for Scientists: https://rmd4sci.njtierney.com/start.html

---
# How do I make that pretty PDF?

- Requires you to install `{tinytex}` - minimal version of LaTeX

```
install.packages("tinytex")
tinytex::install_tinytex()
```

- Yihui's rule of thumb: knit to HTML first, then word or pdf as needed.


---
class: center, middle
# Some YAML tips

---
# Rule 1: Front matter goes first

You can't put code chunks or anything in front of the front matter!

Which is probably why it's called *front matter*.

---
# Rule 2: Be Careful with your spacing

YAML can only use spaces, not tabs. This is an easy mistake to make!

Each "level" uses two spaces, not a tab.

If your YAML is giving you an error, this is usually the reason.

---
# Rule 3: When in doubt, use quotes or `|` for long entries

If you have a long bit of text you want to put in a field, don't hesitate to put it in quotes:

```
author: "My Name is Extremely Long but that's ok"
```

You can also use a `|` on the first line to split text up on multiple lines:

```
author: |
  This is a multi line
  example and you can break 
  anywhere
```

---
# Rule 4: Not all outputs have the same options

- `html_document` is the most flexible (can also embed interactive visualizations)
- `distill::distill_article` has the most intelligent defaults for scientific publishing
- `pdf_document` is probably the least flexible because LaTeX is very picky.

When in doubt, look at the appropriate chapter in the RMarkdown Book. 

---
# Rule 5: Use `{ymlthis}` to write your YAML

Writing YAML can be frustrating! The `ymlthis` package is probably the most error free way to write YAML.

.pull-left[
```{r eval=FALSE}
library(ymlthis)

yml() %>%
  yml_author("Ted Laderas") %>%
  yml_title("My new document") %>%
  yml_subtitle("Which is so great") %>%
  yml_output(html_document(
    toc = TRUE, 
    number_sections = TRUE))
```
]

.pull-right[
```{r echo=FALSE, warning=FALSE}
library(ymlthis)

yml() %>%
  yml_author("Ted Laderas") %>%
  yml_title("My new document") %>%
  yml_subtitle("Which is so great") %>%
  yml_output(html_document(toc = TRUE, number_sections = TRUE))
```
- can cut and paste this into your new document
]


---
# Resources for Further Learning

- [RMarkdown Cheatsheet](https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)
- [RMarkdown Cookbook](https://bookdown.org/yihui/rmarkdown-cookbook/)
- [RMarkdown for Scientists](https://rmd4sci.njtierney.com/)
