# Part 3: More Data Visualization and Data Manipulation


## Learning Objectives

- **Create** and **modify** scatterplots and boxplots
- **Split** figures into multiple panels using `facet_wrap()`
- **Customize** your plots using built in `theme()`s
- **Sort** by a variable in a dataset using `arrange()`
- **Select** variables in a dataset using `select()`
- **Filter** a dataset using the `filter()` function



## Getting set up

```{r message=FALSE, warning=FALSE}
# load library
library(tidyverse)
library(readxl)
library(here)

smoke_complete <- read_excel(here("data/smoke_complete.xlsx"), 
                             sheet=1, 
                             na="NA")


#remove some columns - we'll talk about this later
smoke_complete <- smoke_complete %>% 
  select(age_at_diagnosis, tumor_stage,
         cigarettes_per_day, gender, 
         vital_status, disease)
```


## More about `ggplot2`

### Customizing a Scatterplot

Now that we have the data generally displayed the way we'd like, we can start to customize a plot.

```{r}
our_plot <- ggplot(smoke_complete) +

  aes(x = age_at_diagnosis, 
      y = cigarettes_per_day, 
      color = disease) +
  
  geom_point() +
  
  labs(title = "Cigarettes per Day versus Age at Diagnosis",
       x = "Age at Diagnosis",
       y = "Cigarettes Smoked per Day")
  
our_plot
```



### Changing visual properties using built in themes

```{r}
our_plot +
  theme_minimal()
```

A complete list of pre-set themes is available [here](https://ggplot2.tidyverse.org/reference/ggtheme.html), 
and we'll cover ways to customize our own themes later in this lesson.



### Using `theme()` to customize

Adding the `theme()` function will let us customize our plot further.

There are a few arguments that are really helpful to modify:

- `axis.title`
- `axis.title.x` 
- `axis.title.y` 
- (The labels for the axes)
- `legend.position` (Placing the legend, including removing it)



```{r}
our_plot + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90))
```


```{r}
our_plot + 
  theme_bw() +
  theme(legend.position = "none")
```



### Saving your work


After you're satisfied with a plot, it's likely you'd want to share it with other people or include in a manuscript or report.

```{r}
# save plot to file
ggsave("awesomePlot.jpg", width = 10, height = 10, dpi = 300)
```


This automatically saves the last plot for which code was executed. This command interprets the file format for export using the file suffix you specify. The other arguments dictate the size (`width` and `height`) and resolution (`dpi`).




## Brief Aside: Categorical Data (`factor`s)


One data type that we haven't yet looked at are `factor`s - factors are how R represents *categorical data*.

For the most part, you can use `character` and `factors` interchangably for categorical data.

However, there is one main difference. `factor`s define the permissible values in a vector with an argument called `levels`. They also define the order in which these values are displayed.


```{r}
character_vector <- c("Dog", "Dog", "Cat", "Mouse")
table(character_vector)
```

You can make a character vector into a factor vector by using the `factor()` function, and supplying an argument called `levels`.


### Levels of a Factor

The `levels` of a factor are the permissible values in a `factor`. The order of the values in a factor also control the order in which the values appear in tables and on the axes in a plot. 

So, you can control the order of the categories in a factor by specifying the order of the categories in the `levels` argument.

```{r}
factor_vector <- factor(character_vector, 
                        levels = c("Dog", "Cat", "Mouse"))

table(factor_vector)
```

This ordering is the main reason to use `factors`. We will revisit this again in Part 4 when we learn how to manipulate the data type of a variable in a `data.frame`.




## Back to `ggplot2`: Boxplots

Boxplots compare the distribution of a quantitative variable among categories.

Remember, `vital_status` is a `character` vector, but we're not too worried about the implicit order of the categories, so we can use it as is in our boxplot.

```{r}
# creating a boxplot

ggplot(smoke_complete) +
  
  aes(x = vital_status, y = cigarettes_per_day) +
  
  geom_boxplot()
```


The main differences from the scatterplots we created earlier are the `geom` type and the variables plotted.

We can change the color similarly to scatterplots. However, we map to `fill` and not `color`:


```{r}
# adding color
ggplot(smoke_complete) +
  
  aes(x = vital_status, 
      y = cigarettes_per_day, 
      fill = vital_status) +
  
  geom_boxplot()
```




### Faceting our boxplot

One of the most powerful ways to change a visualization is by *faceting*. We can make multiple plots using another categorical variable.

To do this, we have to add the `facet_wrap()` command to our plot. We need to specify the variable to `facet_wrap` - `disease` by using the `vars()` function to specify it as a variable.


```{r}
# adding color
ggplot(smoke_complete) +
  
  aes(x = vital_status, 
      y = cigarettes_per_day, 
      fill = vital_status) +
  
  geom_boxplot() +
  
  ylim(c(0,20)) +
  
  facet_wrap(vars(disease))
```


Don't forget to look at the help documentation (e.g., `?facet_wrap`) to learn more about additional ways to customize your plots!


## Further learning

If you are interested in learning more about ggplot:
- Documentation for all `ggplot` features is available [here](https://ggplot2.tidyverse.org).
- RStudio also publishes a [ggplot cheat sheet](https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf) that is really handy!





## Data Manipulation using `dplyr`

What we have learned is pretty great overall for plotting. But there are a lot of other things we can do with `data.frame`s!

We're going to talk about another tidyverse package: `dplyr`. `dplyr` is your all-purpose toolbox for filtering, summarizing, and transforming data.

The nice thing about `dplyr` is that it is focused on verbs that do things to your data.


We'll focus on the following verbs:

- `arrange()` - sorting a `data.frame` by a variable
- `filter()` - subsetting a `data.frame` by criteria
- `select()` - returning only a few columns from our `data.frame`
- `group_by()/summarize()` - summarizing our data.frame, such as counting and computing means
- `mutate()` - transforming variables in our data
- `%>%` - the pipe character, which lets us join our verbs together in a *pipeline*.




### Introducing the pipe (`%>%`)


Often times, we want to do multiple operations on our data and in a specific order.

For example, I might want to do the following:


*Take* my dataset `smoke_complete` **and then**
*Sort* it by `cigarettes_per_day` **and then**
*filter* to have only `males` from the data.

The pipe (`%>%`) function acts like the **and then**:


```{r}
# Take my dataset smoke complete **and then**
smoke_complete %>%
  
#sort it by cigarettes_per day **and then**
  arrange(cigarettes_per_day) %>%

#filter it to only have males
  filter(gender == "male")

```


You can think of a pipe as putting the output of one step as an input of another. These two statements are equivalent:


```{r}
smoke_complete %>%
  arrange(cigarettes_per_day)
```

and

```{r}
arrange(smoke_complete, cigarettes_per_day)
```


One big advantage of the pipe is that you can build your processing line by line.

In practice, as I work, I will often pipe things into `View()` to confirm I did things correctly:


```{r eval = FALSE}
smoke_complete %>%
  arrange(cigarettes_per_day) %>%
  View()

```


### The difference between `+` and `%>%`

Remember that `+` is for `ggplot2` and that `%>%` is for `dplyr`. To keep them distinct and avoid confusion, separate our your data processing and your plotting:


```{r}
arranged_smokers <- smoke_complete %>%
  filter(gender == "male")

ggplot(data=arranged_smokers) +
  aes(x = cigarettes_per_day) +
  geom_histogram()
```




### Sorting Data Frames using `arrange()`


Ok, we already have encountered `arrange()`. It's a function that lets us sort a `data.frame` by a variable.

By default, it sorts in *ascending* order:

```{r}
smoke_complete %>% 
  arrange(cigarettes_per_day)
```


To sort by descending order, you need to wrap the variable in the `desc()` function:

```{r}
smoke_complete %>%
  arrange(desc(cigarettes_per_day))
```


You can also arrange by multiple variables. 

```{r}
smoke_complete %>%
  arrange(desc(cigarettes_per_day),
          tumor_stage)
```


Note that order of variables in `arrange()` matters!

```{r}
smoke_complete %>%
  arrange(
    tumor_stage,
    desc(cigarettes_per_day)
          )
```



### `filter()`ing our data


`filter()` is an extremely powerful function. It lets us subset our data according to specific criteria.

Let's filter on a `numeric` variable, `cigarettes_per_day`:


```{r}
smoke_complete %>%
  filter(cigarettes_per_day < 20)

```


We can also filter on a category. But that does require us to know the values of that categorical variable.


```{r}
smoke_complete %>%
  filter(tumor_stage == "stage iv")
```




### Filtering requires a little logic

We can chain multiple criteria using the `&` (AND) or `|` (OR) operators. But we need to review a little logic before we do this.

If I wanted to return patients who were 

`male` **and** `stage iv`, 

I would want use an `&` to chain these criteria together:

```
gender == "male" & tumor_stage == "stage_iv"
```

```{r}
smoke_complete %>%
  filter(gender == "male" & 
           tumor_stage == "stage iv")

```


If I wanted patients who were

`male` **or** `stage iv`

I would use an `|` to chain these criteria together.

```{r}
smoke_complete %>%
  filter(gender == "male" |
           tumor_stage == "stage iv")
```


Think about it: which of the above two code blocks will return a larger number of patients?


### More about Comparison and Logical Operators


This is a useful reference for all the different operators (both logical and comparison) that you can use: https://www.datamentor.io/r-programming/operator/



### Selecting columns using `select()`

The final verb we'll look at is `select()`. It allows us to select variables from our dataset:


```{r}
smoke_complete %>%
  select(gender, tumor_stage)

```


If we want to select everything *but* one variable, we can use a `-` in front of that variable.

```{r}
smoke_complete %>%
  select(-gender)

```



### The difference between `filter()` and `select()`


One thing to keep in mind is that:

`filter()` works on rows, and
`select()` works on columns

Keep that in mind!



### Saving our results

Let's save our processed data in the `data/` directory. We'll save it as a `csv` file, which is short for *comma separated value*. This is a file type that can be easily imported into excel.


```{r}
processed_data <- smoke_complete %>%
  select(-gender) %>%
  filter(cigarettes_per_day < 20)

write_excel_csv(x = processed_data,
                path = "data/processed_data.csv")
```


## What you learned today

- customizing `ggplots` using `theme()` 
- making boxplots 
- faceting plots
- a little bit about factors
- pipes
- `arrange()`
- `filter()`
- `select()`



## Practice

Try out chapters 2 and 3 in the R-Bootcamp:

https://r-bootcamp.netlify.app/chapter2 
https://r-bootcamp.netlify.app/chapter3


## Acknowledgements

This notebook was adapted from material from Kate Hertweck and https://fredhutch.io and from the R-Bootcamp by Ted Laderas and Jessica Minnier.



## Assignment 3

1. Load the `smoke_complete` dataset from the `data` folder. Use `filter()` to subset those patients who have 

disease == "LUSC"

then save the results to an object called `lusc_smokers`:

```{r}
smoke_complete <- read_excel("data/smoke_complete.xlsx", sheet =1, na= "NA")


```

2. Sort `lusc_smokers` by decreasing `tumor_stage`:

```{r}


```


3. For `lusc_smokers`, make a boxplot where y = cigarettes_per_day and x = tumor_stage.


```{r}


```

3. Modify your plot to facet by `gender`:


```{r}


```


4. Subset `smoke_complete` to a different set of patients, such as gender == "female".

Make a new faceted plot, this time a scatterplot. You're free to choose two numeric variables of interest to plot from `smoke_complete`. 

Color your plot by `disease` and facet by `vital_status`.


```{r}




```

